---
title: "Controlling misclassification bias in longitudinal udder health studies"
author: "Denis Haine"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Controlling misclassification bias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: dufour2012
  title: Epidemiology of coagulase-negative staphylococci intramammary infection in dairy cattle and the effect of bacteriological culture misclassification
  author:
  - family: Dufour
    given: Simon
  - family: Dohoo
    given: Ian
  - family: Barkema
    given: Herman
  - family: DesCôteaux
    given: Luc
  - family: DeVries
    given: Trevor
  - family: Reyher
    given: Kristen
  - family: Scholl
    given: Daniel
  container-title: Journal of Dairy Sciences
  volume: 95
  issue: 6
  page: 3110-3124
  type: article-journal
  issued:
    year: 2012
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Using imperfect tests may lead to biased estimates of disease frequency and of
associations between risk factors and disease.
For instance in longitudinal udder health studies, both quarters at risk and
incident intramammary infections (IMI) can be wrongly identified, resulting in
selection and misclassification bias, respectively.
Diagnostic accuracy can possibly be improved by using duplicate or triplicate
samples for identifying quarters at risk and, subsequently, incident IMI.

This package allows the udder health researcher to estimate the potential biases
that would be present in his/her study according to pathogen, population, and test
characteristics.
Improvement in sensitivity (Se) and/or specificity (Sp) of the tests can also be
assessed with the package to help the design and planning of the study.
The package was developed for the paper [Diagnosing intramammary infection:
Controlling misclassification bias in longitudinal udder health
studies](https://doi.org/10.1016/j.prevetmed.2017.11.010), by D. Haine, I.
Dohoo, D. Scholl, and S. Dufour.
Full description of the methodology is provided in the above paper.
The approach in the paper used many simulations to explore various biases (i.e.
had to resort to cloud computing for these lengthy simulations), but the
researcher interested in assessing his/her study design can use the package with
less simulations to have an indication on the best planning for his/her study.
This is what is presented hereafter.

The package is installed from Github:
```{r, eval=FALSE}
devtools::install_github('dhaine/misclass')
```

It is loaded into your session with:
```{r}
library(misclass)
```

The package uses prior Beta distributions for sensitivity and specificity of the
diagnostic test(s), and logistic and Poisson 3-level (quarters within cows
within herds) regression models for evaluating cluster-specific association
(odds ratio) and incidence.
The models are implemented using the [Stan](http://mc-stan.org/) modelling
language.

## Creating an hypothetical cohort study

The function `make_data` allows the creation of datasets simulating an
hypothetical cohort study.
Since correlation structures are often found in udder health studies, the
generated datasets are deemed to have occurred from the collection of two milk
samples collected 1 month apart from each quarters of a random sample of cows
per herd, from a given number of dairy herds.
The first milk sample (S~1~) is used to identify quarters at risk of intramammary
infection (IMI) at the beginning of the cohort, while the second (S~2~) is used to
identify the outcome (acquisition of a new IMI).
Three hypothetical exposures `E_q`, `E_c`, `E_h` (quarter, cow, and herd level)
with known strength of association are generated.
As it is often the case [@dufour2012], exposures are equally associated with
odds of a prevalent IMI on first milk sample as with odds of IMI acquisition on
second sample.
Exposures are randomly associated with odds of eliminating an existing IMI (OR =
1.0).
If *S. aureus* or CNS (coagulase-negative staphylococci) is chosen, default
parameters are used.
Otherwise user has to provide his/her own.

So let's say our researcher wants to run a longitudinal study where he/she will
recruit 20 herds, and within each herd 20 cows will be sampled.

```{r}
sim_list <- vector("list", 10)
set.seed(123)
sim_list <- replicate(n = 10, expr = make_data(20, 20, "cns"), simplify = FALSE)
```
In the above code, first five empty lists are created that will hold the
simulated datasets.
Then a seed is set for reproducibility.
Then the `make_data` function is used to create a dataset made of 20 herds, 20
cows per herd, with the default parameter values for CNS (see `?make_data` for
setting parameters).
This is replicated 10 times to allow for 10 simulations later on.

Increasing Se and/or Sp is commonly achieved in udder health studies by carrying
on duplicate or triplicate samplings with parallel or series interpretation.
The gains or losses in Se/Sp are added/substracted to/from the original
distributions for Se and Sp, and biased S~1~' and S~2~' variables are generated.
Note that you can replace duplicate/triplicate samplings by whatever value you
hope a different test would bring to your study.

## Checking presence of bias for association measure

The `check_bias` function fit a Bayesian logistic regression model to evaluate
the true association, and the association in presence of selection and/or
misclassification bias.

```{r, warning=FALSE, message=FALSE, results='hide'}
bias_or <- check_bias(sim_list,
                      iter = 250,
                      warmup = 50,
                      chains = 1,
                      cores = 1,
                      seed = 123,
                      nsimul = 10)
```

The object created, `bias_or` is a list holding the results for the 10
simulations with, for each simulation, the values for the incidence (the first 5
elements of the list, as we ran 5 simulations), i.e. true association, single
sample, selection bias only, and misclassification bias only, and diagnostic
parameters in the next 5 elements of the list (Rhat, n_eff, etc.).
You can use `str(bias_or)` to check the structure of this list.

We can put each together into separate matrices:

```{r}
diag <- do.call(rbind, bias_or[11:20])
bias_check <- do.call(rbind, bias_or[1:10])
```

### Diagnostics

You can run several diagnostics checks.
More info on diagnostics can be found on the [Stan
wiki](https://github.com/stan-dev/stan/wiki/Stan-Best-Practices) page.

```{r, fig.show='hold'}
## Distribution of Rhat 
hist(diag[, 1])
## Ratio n_eff/sample_size
hist(diag[, 2] / diag[, 3])
```

```{r}
Rh <- diag[, 1]
Rh[Rh > 1.1]
summary(diag)
```

## Assess bias

A summary of the different OR can be obtained:

```{r}
summary(bias_check)

exp(median(bias_check[, 1]))
exp(median(bias_check[, 4]))
```

We see that the misclassification bias move the association towards the null
value (OR of `r round(exp(median(bias_check[, 4])), 2)` versus true OR of 
`r round(exp(median(bias_check[, 1])), 2)`).

We can generate a density plot to visualize the posterior distributions.
But first let's write a function to "densify" the data:

```{r}
## Densify data to be used in a density plot 
densify <- function(df, simul) {
    ncolumns <- ncol(df)
    dflist <- list()
    CrI <- matrix(NA, nrow = ncolumns, ncol = 3)
    colnames(CrI) <- c("q2.5", "q50", "q97.5")
    for (i in 1:ncolumns) {
        pre.dens <- reshape2::melt(df[, i])
        pre.dens <- exp(pre.dens)
        CrI[i, ] <- quantile(df[, i], c(.025, .5, .975))
        dens <- density(pre.dens$value)
        dens <- data.frame(x = dens$x, y = dens$y)
        dens$simul <- simul[i]
        dflist[[i]] <- dens
    }
    densified <- do.call(rbind, dflist)
    out_list <- list(densified, CrI)
    out_list
}

bias_dens <- densify(bias_check,
                     simul = c("True association",
                               "Single sample",
                               "Selection bias only",
                               "Misclassificiation bias only"))
```

```{r, fig.width=6, fig.height=4}
library(ggplot2)
ggplot(bias_dens[[1]], aes(x, y, colour = simul)) +
    geom_line() +
    xlab("Odds ratio") +
    ylab("Density") +
    coord_cartesian(xlim = c(0, 6)) +
    scale_x_continuous(breaks = seq(0, 6, 1)) +
    guides(colour = guide_legend(title = NULL)) +
    theme(legend.position = c(.7, .7))
```

## Checking the effect of different sampling strategies

The same can be done for evaluating the effect of duplicate or triplicate
samplings (i.e., modifying the Se/Sp).
We're not running the code but it would be:

```{r, eval=FALSE}
stgy_or <- sample_stgy(sim_list,
                       iter = 250,
                       warmup = 50,
                       chains = 1,
                       cores = 1,
                       seed = 123,
                       nsimul = 10)
```

## Measure of incidence

Instead of using `check_bias` and `smply_stgy` functions, you use
`check_incidence` and `sample_incidence`, respectively.

Note that Poisson models are slower to converge.

## References
